\chapter{Basic Concepts}
\label{Chp:Concepts}

The GraphBLAS C API is used to construct  
graph algorithms expressed ``in the language of linear algebra.''
Graphs are expressed as matrices, and the operations over 
these matrices are generalized through the use of a
semiring algebraic structure.

In this chapter, we will define the basic concepts used to
define the GraphBLAS C API.  We provide the following elements:
\begin{itemize}
\item Glossary of terms used in this document.  

\item Notation

\item Execution model

\item Error model

\end{itemize}

\section{Glossary}

%TGM I'm leaving a few definitions in here just as examples

\subsection{Basic definitions}

\glossBegin

\glossItem{application} A program that calls methods from the GraphBLAS C API to
solve a problem.

\glossItem{GraphBLAS C API} The application programming interface that fully defines the types, objects, 
literals, and other elements of the C binding to the GraphBLAS.

\glossEnd

\subsection{Objects and their structure}

\glossBegin
\glossItem{handle}  A variable that uses one of the GraphBLAS opaque data types.
The value of this variable holds a reference to a GraphBLAS object but not the contents of the object itself.
Hence, assigning a value of one handle to another variable copies the reference to the GraphBLAS object
but not the contents of the object.

\glossItem{non-opaque datatype} Any datatype that exposes its internal structure.   
This is contrasted
with an \emph{opaque datatype} that hides its internal structure and can
be manipulated only through an API.

\glossEnd



\vfill

\newgeometry{left=2.5cm,top=2cm,bottom=2cm}

\section{Notation}

\begin{tabular}[H]{l|p{5in}}
Notation & Description \\
\hline
$\Dout, \Dinn, \Din1, \Din2$  & Refers to output and input domains of various GraphBLAS operators. \\
$\bDout(*), \bDinn(*),$ & Evaluates to output and input domains of GraphBLAS operators (usually \\
~~~~$\bDin1(*), \bDin2(*)$ & a unary or binary operator, or semiring). \\
$\mathbf{D}(*)$   & Evaluates to the (only) domain of a GraphBLAS object (usually a monoid, vector, or matrix). \\ 
$f$             & An arbitrary unary function, usually a component of a unary operator. \\
$\mathbf{f}(F_u)$ & Evaluates to the unary function contained in the unary operator given as the argument. \\
$\odot$         & An arbitrary binary function, usually a component of a binary operator. \\
$\mathbf{\bigodot}(*)$ & Evaluates to the binary function contained in the binary operator or monoid given as the argument. \\
$\otimes$       & Multiplicative binary operator of a semiring. \\
$\oplus$        & Additive binary operator of a semiring. \\
$\mathbf{\bigotimes}(S)$ & Evaluates to the multiplicative binary operator of the semiring given as the argument. \\
$\mathbf{\bigoplus}(S)$ & Evaluates to the additive binary operator of the semiring given as the argument. \\
$\mathbf{0}(*)$   & The identity of a monoid, or the additive identity of a GraphBLAS semiring. \\
$\mathbf{L}(*)$   & The contents (all stored values) of the vector or matrix GraphBLAS objects.  For a vector, it is the set of (index, value) pairs, and for a matrix it is the set of (row, col, value) triples. \\
$\mathbf{v}(i)$ or $v_i$   & The $i^{th}$ element of the vector $\vector{v}$.\\
$\mathbf{size}(\vector{v})$ & The size of the vector $\vector{v}$.\\
$\mathbf{ind}(\vector{v})$ & The set of indices corresponding to the stored values of the vector $\vector{v}$.\\
$\mathbf{nrows}(\vector{A})$ & The number of rows in the $\matrix{A}$.\\
$\mathbf{ncols}(\vector{A})$ & The number of columns in the $\matrix{A}$.\\
$\mathbf{indrow}(\vector{A})$ & The set of row indices corresponding to rows in $\matrix{A}$ that have stored values.  \\
$\mathbf{indcol}(\vector{A})$ & The set of column indices corresponding to columns in $\matrix{A}$ that have stored values. \\
$\mathbf{ind}(\vector{A})$ & The set of $(i,j)$ indices corresponding to the stored values of the matrix. \\
$\mathbf{A}(i,j)$ or $A_{ij}$ & The element of $\matrix{A}$ with row index $i$ and column index $j$.\\
$\matrix{A}(:,j)$ & The $j^{th}$ column of the the matrix $\matrix{A}$.\\
$\matrix{A}(i,:)$ & The $i^{th}$ row of the the matrix $\matrix{A}$.\\
$\matrix{A}^T$ &The transpose of the matrix $\matrix{A}$. \\
$\neg\matrix{M}$ & The complement of $\matrix{M}$.\\
$\vector{\widetilde{t}}$ & A temporary object created  by the GraphBLAS implementation. \\
$<type>$ & A method argument type that is {\sf void *} or one of the types from Table~\ref{Tab:PredefinedTypes}. \\
{\sf GrB\_ALL} & A method argument literal to indicate that all indices of an input array should be used.\\
{\sf GrB\_Type} & A method argument type that is either a user defined type or one of the  types from Table~\ref{Tab:PredefinedTypes}.\\
{\sf GrB\_Object} &  A method argument type referencing any of the GraphBLAS object types.\\
{\sf GrB\_NULL} & The GraphBLAS NULL.\\
\end{tabular}

\restoregeometry

\section{Error model}



\section{Execution Model}
\label{Sec:ExecutionModel}

%% TGM: I've left the GraphBLAS model here just as a point of reference.

A program using the GraphBLAS C API constructs GraphBLAS objects,
manipulates them to implement a graph algorithm, and then extracts
values from the GraphBLAS objects as the result of the algorithm.
Functions defined within the GraphBLAS C API that manipulate GraphBLAS
objects are called \emph{methods}.  If the method corresponds to one
of the operations defined in the GraphBLAS mathematical specification,
we refer to the method as an \emph{operation}.

Graph algorithms are expressed as an ordered collection of GraphBLAS
method calls defined by the order they are encountered in a program.
This is called the \emph{program order}.  Each method in the collection
uniquely and unambiguously defines the output GraphBLAS objects based
on the GraphBLAS operation and the input GraphBLAS objects. This is the
case as long as there are no execution errors, which can put objects in
an invalid state (see Section~\ref{Sec:ErrorModel}).

The GraphBLAS method calls in program order are organized into contiguous
and nonoverlapping \emph{sequences}.  A sequence is an ordered collection
of method calls as encountered by an executing thread. (For more on
threads and GraphBLAS, see Section~\ref{Sec:ThreadSafety}.)  A sequence
begins with either (1) the first GraphBLAS method called by a thread,
or (2) the first method called by a thread after the end of the
previous sequence.  A sequence can end (terminate) in a variety of ways.
A call to the GraphBLAS {\sf GrB\_wait()} method (Section~\ref{Sec:GrB_wait})
always ends a sequence.  The GraphBLAS {\sf GrB\_finalize()} method
(Section~\ref{Sec:GrB_finalize}) also implicitly ends a sequence. Finally,
in blocking mode (see below), each GraphBLAS method starts and ends its
own sequence.

The GraphBLAS objects are fully defined at any point in a sequence by
the methods in the sequence as long as there are no execution errors.
In particular, as soon as a GraphBLAS method call returns, its output
can be used in the next GraphBLAS method call.  However, individual
operations in a sequence may not be \emph{complete}. We say that an
operation is complete when all the computations in the operation have
finished and all the values of its output object have been produced and
committed to the address space of the program. Furthermore, no additional
execution time can be charged to a completed operation and no additional
errors can be attributed to a completed operation.

The opaqueness of GraphBLAS objects allows execution to proceed
from one method to the next even when operations are not complete.
Processing of nonopaque objects is never deferred in GraphBLAS. That is,
methods that consume nonopaque objects (\eg, {\sf GrB\_Matrix\_build()},
Section~\ref{Sec:Matrix_build}) and methods that produce nonopaque objects (\eg,
{\sf GrB\_Matrix\_extractTuples()}, Section~\ref{Sec:Matrix_extractTuples})
always finish consuming or producing those nonopaque objects before
returning.   

\comment{Furthermore, methods that extract values from opaque GraphBLAS objects
into nonopaque user objects (see Table~\ref{Tab:ExtractMethods})
always force completion of all pending computations on the 
corresponding GraphBLAS source object.

\begin{table}[htb]
    \hrule
    \begin{center}
        \caption{Methods that extract values from a GraphBLAS object, thereby
        forcing completion of the operations contributing to that particular object.}
        \label{Tab:ExtractMethods}

        \begin{tabular}{l|l}
            Method    & Section \\ \hline

            {\sf GrB\_Vector\_nvals}        & \ref{Sec:Vector_nvals}        \\
            {\sf GrB\_Vector\_extractElement}     & \ref{Sec:extract_single_element_vec}    \\
            {\sf GrB\_Vector\_extractTuples}    & \ref{Sec:Vector_extractTuples}    \\
            {\sf GrB\_Matrix\_nvals}        & \ref{Sec:Matrix_nvals}        \\
            {\sf GrB\_Matrix\_extractElement}     & \ref{Sec:extract_single_element_mat}    \\
            {\sf GrB\_Matrix\_extractTuples}    & \ref{Sec:Matrix_extractTuples}    \\
            {\sf GrB\_reduce} (vector-scalar variant)        & \ref{Sec:Reduce_vector_scalar}        \\
            {\sf GrB\_reduce} (matrix-scalar variant)        & \ref{Sec:Reduce_matrix_scalar}        \\
        \end{tabular}
    \end{center}
    \hrule
\end{table}
}


\section{Error Model}
\label{Sec:ErrorModel}

%%TGM  I've left the GraphBLAS model here as a point of reference

All GraphBLAS methods return a value of type {\sf GrB\_Info} to provide
information available to the system at the time the method returns. The
returned value can be either {\sf GrB\_SUCCESS} or one of the defined
error values shown in Table~\ref{Tab:ErrorValues}. The errors fall into
two groups: API errors (Table~\ref{Tab:ErrorValues}(a)) and execution
errors (Table~\ref{Tab:ErrorValues}(b)).

\begin{table}[bh]
\hrule
\begin{center}
\caption{Error values returned by GraphBLAS methods.}
\label{Tab:ErrorValues}

\vspace{1\baselineskip}
(a) API errors
\vspace{1\baselineskip}

\begin{tabular}{l|p{3in}}
Error code    & Description \\ \hline
{\sf GrB\_UNINITIALIZED\_OBJECT} & A GraphBLAS object is passed to a method before {\sf new} was called on it.\\
{\sf GrB\_NULL\_POINTER} & A NULL is passed for a pointer parameter. \\
{\sf GrB\_INVALID\_VALUE} & Miscellaneous incorrect values. \\
{\sf GrB\_INVALID\_INDEX} & Indices passed are larger than dimensions of the matrix or vector being accessed. \\
{\sf GrB\_DOMAIN\_MISMATCH} & A mismatch between domains of collections and operations when user-defined domains are in use.\\
{\sf GrB\_DIMENSION\_MISMATCH} & Operations on matrices and vectors with incompatible dimensions. \\
{\sf GrB\_OUTPUT\_NOT\_EMPTY} & An attempt was made to build a matrix or vector using an output object that already contains valid tuples (elements).\\
%{\sf GrB\_NO\_VALUE} & An attempt was made to extract a value from a tuple within a matrix or vector for which there is no stored value. 
{\sf GrB\_NO\_VALUE} & A location in a matrix or vector is being accessed that has no stored value at the specified location. \scott{It depends on whether or not the non-opaque scalar is
well-defined on return from {\sf extract}}\\
\end{tabular}

\vspace{1\baselineskip}
(b) Execution errors
\vspace{1\baselineskip}

\begin{tabular}{l|p{3in}}
Error code    & Description \\ \hline
{\sf GrB\_OUT\_OF\_MEMORY}         & Not enough memory for operations. \\
{\sf GrB\_INSUFFICIENT\_SPACE}     & The array provided is not large enough to hold output. \\
{\sf GrB\_INVALID\_OBJECT}         & One of the opaque GraphBLAS objects (input or output) is in an invalid state caused by a previous execution error. \\
{\sf GrB\_INDEX\_OUT\_OF\_BOUNDS}  & Reference to a vector or matrix element that is outside the defined dimensions of the object. \\
{\sf GrB\_PANIC}        & Unknown internal error. \\
\end{tabular}

\end{center}
\hrule
\end{table}

An API error means that a GraphBLAS method was called with parameters that
violate the rules for that method.  These errors are restricted to those
that can be determined by inspecting the types and domains of GraphBLAS
objects, GraphBLAS operators, or the values of scalar parameters fixed at
the time a method is called.  API errors are deterministic and consistent
across platforms and implementations.  API errors are never deferred,
even in nonblocking mode. That is, if a method is called in a manner
that would generate an API error, it always returns with the appropriate
API error value.  If a GraphBLAS method returns with an API error, it
is guaranteed that none of the arguments to the method (or any other
program data) have been modified.

Execution errors indicate that something went wrong during the execution
of a legal GraphBLAS method invocation.  Their occurrence may depend on
specifics of the executing environment and data values being manipulated.
This does not mean that execution errors are the fault of the GraphBLAS
implementation.  For example, a memory leak could arise from an error in
an application's source code (a ``program error''), but it may manifest
itself in different points of a program's execution (or not at all)
depending on the platform, problem size, or what else is running at
that time.  Index-out-of-bounds and insuficient space execution errors
always indicate a program error.

In blocking mode, where each method executes to completion, a returned
execution error value applies to the specific method.  If a GraphBLAS
method, executing in blocking mode, returns with any execution error
from Table~\ref{Tab:ErrorValues}(b) other than {\sf GrB\_PANIC}, it
is guaranteed that no argument used as input-only has been modified.
Output arguments may be left in an invalid state, and their use downstream
in the program flow may cause additional errors.  If a GraphBLAS method
returns with a {\sf GrB\_PANIC} execution error, no guarantees can be
made about the state of any program data.

In nonblocking mode, execution errors can be deferred.  A return value
of {\sf GrB\_SUCCESS} only guarantees that there are no API errors in
the method invocation.  If an execution error value is returned by a
method in nonblocking mode, it indicates that an error was found during
execution of the sequence, up to and including the {\sf GrB\_wait()}
method (Section~\ref{Sec:GrB_wait}) call that ends the sequence. When possible, that return value
will provide information concerning the cause of the error.

As discussed in Section~\ref{Sec:GrB_waitOne}, a {\sf GrB\_wait(obj)} on
a specific GraphBLAS object {\sf obj} does not necessarily end a
sequence. However, no additional errors on the methods of the sequence that 
have {\sf obj} as an {\sf OUT} or {\sf INOUT} argument can be reported.
From a GraphBLAS perspective, those methods are {\em complete}.

If a GraphBLAS method, executing in nonblocking mode, returns with
any execution error from Table~\ref{Tab:ErrorValues}(b) other than
{\sf GrB\_PANIC}, it is guaranteed that no argument used as input-only
through the entire sequence has been modified.  Any output argument in
the sequence may be left in an invalid state and its use downstream in the
program flow may cause additional errors.  If a GraphBLAS method returns
with a {\sf GrB\_PANIC}, no guarantees can be made about the state of
any program data.

\begin{figure}[tb]
    \hrule
    \vspace{1\baselineskip}
    \begin{center}
        \begin{minipage}{3in}
            \begin{verbatim}
            const char *GrB_error();
            \end{verbatim}
        \end{minipage}
    \end{center}
    \caption{Signature of {\sf GrB\_error()} function.}
    \label{Fig:GrB_error}
    \hrule
\end{figure}

After a call to any GraphBLAS method, the program can retrieve additional
error information (beyond the error code returned by the method) though a
call to the function {\sf GrB\_error()}. The signature of that function is
shown in Figure~\ref{Fig:GrB_error}.  The function returns a pointer to a 
NULL-terminated string, and the contents of that string are implementation 
dependent. In particular, a null string (not a {\sf NULL} pointer) is always a valid error string.
The pointer is valid until the next call to any GraphBLAS method by the same thread.
{\sf GrB\_error()} is a thread-safe function, in the sense that multiple threads can
call it simultaneously and each will get its own error string back, referring to the
last GraphBLAS method it called.
